The non-random assembly of functional motifs in plant-pollinator networks worldwide

### 2022-02-08 Manuscript/Main_text_Oikos.Rmd
Specificity plot ready for supplementary

---
title: "The non-random assembly of functional motifs in plant-pollinator networks worldwide"
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
- \usepackage{caption,setspace}
- \captionsetup[figure]{labelformat=empty}
- \floatplacement{figure}{H}
- \usepackage[skip=3pt]{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{lscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
output:
  pdf_document: default
  html_document: default
  word_document: default
always_allow_html: true
bibliography: bibliography.bib
---

@@

Despite motifs are abstract representations often decoupled from species ecology, linking the structural properties of the meso-scale with the species ecology can help the understanding of ecological processes. For instance, different motifs can have different ecological meanings [@simmons2019] and the position within a motif can determine the species functional role [@stouffer2012; @baker2015]. However, it is unclear how the species ecology and life history traits determines the species' functional role within the network of interactions [@coux2016]. For example, large pollinators can forage larger distances [@greenleaf2007], deposit greater pollen quantities [@foldesi2021] and handle complex zygomorphic flowers in comparison with small pollinators that are restricted to lower floral complexity [@gong2009]. How this different pollinator behavior translates into their interaction topology is unknown. Similarly, recent empirical findings indicate that the meso-scale is the best descriptor of plant reproductive success [@allen2021], but little is known about how plants reproductive strategies shape their position within the network of interactions. Although some studies have evaluated plant reproductive strategies in plant-pollinator networks [@tur2013; @lazaro2020], this is an often overlooked aspect in a community context [@devaux2014] and rarely incorporated into plant-pollinator network studies. Hence, exploring how the main plant reproductive strategies integrate with the emergent motif framework can shed light on key aspects of ecosystem functioning.

Here, we used 60 plant-pollinator networks from 18 different studies and 14 countries, alongside functional groups of plants and floral visitors based on functional traits and taxonomic rank, respectively. To obtain plant functional groups, we used a comprehensive dataset that included floral, reproductive and vegetative traits compiled in a larger set of networks. Floral visitor functional groups were divided by the main taxonomic groups that differed in life form and behaviour. Once we split the different plant-pollinator networks into their motif elements, we explored: (i) if there is a common invariant structural property in the overall motif networks (i.e. over- and under-represented motifs); (ii) which functional groups are over- and under-represented in the different motif positions; and, (iii) if there are over- and under-represented functional group combinations within a motif.

@@

Second, floral visitors were grouped into functional groups based on taxonomic rank as done similarly in past plant-pollinator studies [@fenster2004; @ollerton2009]. We opted to divide floral visitors on the taxonomic rank level and not with functional traits because the main orders of floral visitors differed in form and behaviour and had lower superior taxonomical complexity (i.e., floral visitors had 6 orders versus plants that had 38). Thus, this allowed us to group floral visitors into functional groups that represented adequately the main life strategies of the possible pollinators: (i) bees (Hymenoptera-Anthophila), (ii) non-bee Hymenoptera (Hymenoptera-non-Anthophila), (iii) syrphids (Syrphidae-Diptera), (iv) non-syrphids-Diptera, (v) Lepidoptera and (vi) Coleoptera. However, a minor set of species belonged to other functional groups that were considered in analyses but not discussed further because of their low representation in the full set of networks (3.55$\%$ of the total interactions recorded). These functional groups were 'lizards', 'birds' and 'other insects'. This last group was formed by a mix of uncommon insect taxa on the full set of networks.

@@

To assess the significance of the observed frequencies, we created 1,000 simulated networks for each binary network using the _nullmodel_ function and the _vaznull_ model in the _bipartite_ package version _2.14_ [@dormann2009]. Generated networks had the same number of plants and floral visitors, as well as the same connectance of their corresponding empirical networks. After extracting the motif frequencies from the simulated networks, for each motif type and empirical network, we calculated the percentage of simulated networks whose frequencies were smaller than the ones observed, that is, we estimated the percentile of the observed motif frequencies. Motifs whose percentile is close to 0 or 100 are under- or over-represented in the empirical networks, respectively, and thus they cannot be predicted by connectance and the number of species alone. To summarize general patterns across networks, we used an intercept-only linear mixed model (LMM) per motif with the help of the package lmer version _1.1-21_ [@bates2015], where the response variable was the observed motif percentile per network. In these models, we used the study identifiers in **Table S1** as a random intercept. By doing so, we obtained estimates of the average motif frequency, but controling the variation at the study level.

**Over and under-represented functional groups on motif positions**

We calculated which functional groups were over or under-represented in different motif positions by comparing position frequencies of empirical networks with those of their corresponding simulated counterparts. To estimate the position frequencies of each functional group in a given network, we added the frequencies of those species that belong to the group, and then, we normalised the resulting frequencies by dividing the position measure for each group by the total number of times that a group appears in any position within the same motif size class. Then, we calculated the percentile of the observed position frequencies for each group and network, just like we did with motif frequencies. To outline the general patterns of position frequencies across networks and functional groups, we fit a LMM per motif position, where the response variable was the observed position percentile per network. We used the functional group identifier as an explanatory variable and the study identifiers as a random intercept. By adding the functional group estimates to the model intercept, we assessed the average motif frequency, after controlling the variation at the study level. Finally, we visualized with the help of the package _ComplexHeatmap_ version _2.6.2_ [@gu2016] over- and under- representation of plant and pollinator functional groups on the different motif positions.

**Over- and under-represented funtional group combinations of motifs**

@@

**Plant and floral visitor functional groups**

@@

**Over and under-represented functional groups on motif positions**

@@

**Over- and under-represented functional group combinations of motifs**

@@

The probabilities of appearance of each functional group on the different motif positions were insufficient to predict most of the observed network position combinations (**Figure 5**). This implies that not just the representation of network motifs types but also the identity of the elements of their structure is influenced by ecological processes that determine under- and over-representation of the different position combinations. In addition, we found that just 30$\%$ of the possible 53,250 motif combinations are realized. Thus, similarly to the presence of forbidden links at the network level [@olesen2011; @jordano2016], we also found the existence of forbidden motif positions and those represented almost two thirds of the total posible combinations of indirect interactions. This concurs well with the general idea that plant-pollinator networks have a high proportion of non-realized direct interactions [@jordano2003; @bascompte2013], and as a result, also a high proportion of non-realized combinations of indirect interactions. Some of this non-observed interactions are probably due to under-sampling [@dorado2011; @jordano2016]. However, a large fraction of the possible interactions do not take place because of life history constraints such as species co-occurrence, trait matching processes and phenological uncoupling [@olesen2011; @bartomeus2016; @jordano2016]. Despite the advances in the understanding on the ecological processes that influence species pairwise interactions [@bartomeus2016; @peralta2020], we still know little of how these processes determine competitive and facilitative interactions in mutualistic networks. Thus, as suggested by @simmons2020, exploring the proportion of these realized network motifs that are competitive or facilitative is a necessary next step to improve our understanding of the architecture of biodiversity [@bascompte2007].

>>>>

---
title: "The non-random assembly of functional motifs in plant-pollinator networks worldwide"
editor_options:
  chunk_output_type: console
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{xcolor}
- \usepackage{caption,setspace}
- \captionsetup[figure]{labelformat=empty}
- \floatplacement{figure}{H}
- \usepackage[skip=3pt]{caption}
- \captionsetup[figure]{labelformat=empty}
- \usepackage{lscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
output:
  pdf_document: default
  html_document: default
  word_document: default
always_allow_html: true
bibliography: bibliography.bib
biblio-style: apalike
csl: oikos.csl #oikos style from zotero styles

---

@@

Despite motifs are abstract representations often decoupled from species ecology, linking the structural properties of the meso-scale with the species ecology can help the understanding of ecological processes. For instance, different motifs can have different ecological meanings [@simmons2019] and the position within a motif can determine the species functional role [@stouffer2012; @baker2015]. However, it is unclear how the species ecology and life-history traits determines the species' functional role within the network of interactions [@coux2016]. For example, large pollinators can forage larger distances [@greenleaf2007], deposit greater pollen quantities [@foldesi2021] and handle complex zygomorphic flowers in comparison with small pollinators that are restricted to lower floral complexity [@gong2009]. How this different pollinator behavior translates into their interaction topology is unknown. Similarly, recent empirical findings indicate that the meso-scale is the best descriptor of plant reproductive success [@allen2021], but little is known about how plants reproductive strategies shape their position within the network of interactions. Although some studies have evaluated plant reproductive strategies in plant-pollinator networks [@tur2013; @lazaro2020], this is an often overlooked aspect in a community context [@devaux2014] and rarely incorporated into plant-pollinator network studies. Hence, exploring how the main plant reproductive strategies integrate with the emergent motif framework can shed light on key aspects of ecosystem functioning.

Here, we used 60 plant-pollinator networks from 18 different studies and 14 countries, alongside functional groups of plants and taxonomic groups of floral visitors based on functional traits and taxonomic rank, respectively. To obtain plant functional groups, we used a comprehensive dataset that included floral, reproductive and vegetative traits compiled in a larger set of networks. Floral visitor groups were divided by the main taxonomic groups that differed in life form and behaviour. Once we split the different plant-pollinator networks into their motif elements, we explored: (i) if there is a common invariant structural property in the overall motif networks (i.e. over- and under-represented motifs); (ii) which plant and pollinator groups are over- and under-represented in the different motif positions; and, (iii) if there are over- and under-represented plant and pollinator group combinations within a motif.

@@

Second, floral visitors were grouped into taxonomic groups based on taxonomic rank as done similarly in past plant-pollinator studies [@fenster2004; @ollerton2009]. We opted to divide floral visitors on the taxonomic rank level and not with functional traits because the main orders of floral visitors differed in form and behaviour and had lower superior taxonomical complexity (i.e., floral visitors had 6 orders versus plants that had 38). Thus, this allowed us to group floral visitors into groups that represented approximately the main life strategies of the possible pollinators: (i) bees (Hymenoptera-Anthophila), (ii) non-bee Hymenoptera (Hymenoptera-non-Anthophila), (iii) syrphids (Syrphidae-Diptera), (iv) non-syrphids-Diptera, (v) Lepidoptera and (vi) Coleoptera. However, a minor set of species belonged to other taxonomic groups that were considered in analyses but not discussed further because of their low representation in the full set of networks (3.55$\%$ of the total interactions recorded). These taxonomic groups were 'lizards', 'birds' and 'other insects'. This last group was formed by a mix of uncommon insect taxa on the full set of networks.

@@

To assess the significance of the observed frequencies, we created 1,000 simulated networks for each binary network using the _nullmodel_ function and the _vaznull_ model in the _bipartite_ package version _2.14_ [@dormann2009]. Generated networks had the same number of plants and floral visitors, as well as the same connectance of their corresponding empirical networks. **MA FORTUNA, what about degree?**. After extracting the motif frequencies from the simulated networks, for each motif type and empirical network, we calculated the percentage of simulated networks whose frequencies were smaller than the ones observed, that is, we estimated the percentile of the observed motif frequencies. Motifs whose percentile is close to 0 or 100 are under- or over-represented in the empirical networks, respectively, and thus they cannot be predicted by connectance and the number of species alone. To summarize general patterns across networks, we used an intercept-only linear mixed model (LMM) per motif with the help of the package lmer version _1.1-21_ [@bates2015], where the response variable was the observed motif percentile per network. In these models, we used the study identifiers in **Table S1** as a random intercept. By doing so, we obtained estimates of the average motif frequency, but controling the variation at the study level.

**Over and under-represented groups on motif positions**

We calculated which plant and pollinator groups were over or under-represented in different motif positions by comparing position frequencies of empirical networks with those of their corresponding simulated counterparts. To estimate the position frequencies of each functional group in a given network, we added the frequencies of those species that belong to the group, and then, we normalised the resulting frequencies by dividing the position measure for each group by the total number of times that a group appears in any position within the same motif size class. Then, we calculated the percentile of the observed position frequencies for each group and network, just like we did with motif frequencies. To outline the general patterns of position frequencies across networks and functional groups, we fit a LMM per motif position, where the response variable was the observed position percentile per network. We used the functional group identifier as an explanatory variable and the study identifiers as a random intercept. By adding the functional group estimates to the model intercept, we assessed the average motif frequency, after controlling the variation at the study level. Finally, we visualized with the help of the package _ComplexHeatmap_ version _2.6.2_ [@gu2016] over- and under- representation of plant and pollinator functional groups on the different motif positions.

**Over- and under-represented group combinations of motifs**

@@

**Plant and floral visitor groups**

@@

**Over and under-represented groups on motif positions**

@@

**Over- and under-represented group combinations of motifs**

@@

The probabilities of appearance of each functional group on the different motif positions were insufficient to predict most of the observed network position combinations (**Figure 5**). This implies that not just the representation of network motifs types but also the identity of the elements of their structure is influenced by ecological processes that determine under- and over-representation of the different position combinations. In addition, we found that just 30$\%$ of the possible 53,250 motif combinations are realized. Thus, similarly to the presence of forbidden links at the network level [@olesen2011; @jordano2016], we also found the existence of forbidden motif positions and those represented almost two thirds of the total posible combinations of indirect interactions. This concurs well with the general idea that plant-pollinator networks have a high proportion of non-realized direct interactions [@jordano2003; @bascompte2013], and as a result, also a high proportion of non-realized combinations of indirect interactions. Some of this non-observed interactions are probably due to under-sampling [@dorado2011; @jordano2016]. However, a large fraction of the possible interactions do not take place because of life-history constraints such as species co-occurrence, trait matching processes and phenological uncoupling [@olesen2011; @bartomeus2016; @jordano2016]. Despite the advances in the understanding on the ecological processes that influence species pairwise interactions [@bartomeus2016; @peralta2020], we still know little of how these processes determine competitive and facilitative interactions in mutualistic networks. Thus, as suggested by @simmons2020, exploring the proportion of these realized network motifs that are competitive or facilitative is a necessary next step to improve our understanding of the architecture of biodiversity [@bascompte2007].
